\documentclass{article}

\usepackage[margin=1.5cm]{geometry}
\usepackage[colorlinks]{hyperref}
\usepackage{natbib}

\begin{document}

\section*{Essential Criteria}

\subsection*{Qualiﬁcations, experience and knowledge}

\paragraph{Possession of a PhD or equivalent qualiﬁcation in a relevant subject
area}
%
I hold a PhD in Electrical Engineering from the
\href{https://www.usc.edu/}{University of Southern California}, completed
within its \href{https://minghsiehece.usc.edu/groups/sipi/}{Signal and Image
Processing Institute}.

\paragraph{An innovative personal research agenda in NeuroAI, computational
neuroscience, or a A/I closely related ﬁeld.}
%
Please refer to my Research Vision described in the attached
rapelaVisionStmt.pdf.

\paragraph{Evidence of the ability to deliver and publish internationally excellent research.
Publications should demonstrate both methodological rigour (e.g. development or
application of quantitative/mathematical frameworks) and engagement with genuine
biological or neuroscientiﬁc questions, not purely technical contributions to AI/ML alone.}
%
I have followed a non-traditional yet highly rigorous career path uniquely
tailored to the modern demands of computational neuroscience. Following a PhD
and two postdoctoral fellowships at the interface of statistical machine
learning and neuroscience—resulting in key publications in leading journals
such as \textit{Neural Computation}, \textit{Network: Computation in Neural
Systems}, and the \textit{Journal of Vision}—I chose to embed myself deeply
within world-class experimental environments. I joined the Gatsby Computational
Neuroscience Unit and the Sainsbury Wellcome Centre (SWC) as a Research
Engineer Fellow to directly bridge the gap between advanced statistical
methodology and world-class neuroscience experimentation.

At the Gatsby Unit and SWC I have grown enormously. I have learned world class
machine learning at the Unit, and meeting one-on-one weekly with the Unit's
director, Maneesh Sahani, for more than seven years. I have acquired
professional software development skills. I have been part of a unique
computational/experimental neuroscience collaborative environment, where I
actively participated in the institute-wide team science Aeon project. I was
awarded £800.000 BBSRC award, and wrote a joint application between the Gatsby
Unit, the SWC and the Allen Institute for Neural Dynamics for a BBSRC/NSF-Bio
award. I lead a small team of research software engineers for more than three
years. I designed and taught the Statistical Neuroscience course for SWC PhD
students for two years. And I mentored two undergraduate students in
one-year-long fellowships sponsored by the Simmons Foundation.

However, during this time my focus was on software development and ML training
and not on research publications.
Now I want to center my academic activities around publishing research
articles, teaching and mentoring students, and collaborating with experimental
groups at NPP and CDB.
%
I am excellently prepared to do this. In the last seven years I have
experimented with several paper ideas that will allow me
to publish several high-quality papers in the next few years. Below I summarise
a few of them:

\begin{itemize}

    \item\textbf{Traveling waves during rhythmic speech production:} Expanding on
        single-subject findings reported in my earlier work (Rapela 2016, 2017,
        2018), I am leveraging a newly acquired dataset across six subjects to
        publish a robust, high-impact characterization of speech-related
        cortical traveling waves.

    \item\textbf{svGPFA distribution and scientific papers:} I am preparing a
        pair of manuscripts documenting the unique numerical and software
        engineering architecture of our Python distribution of svGPFA,
        alongside a scientific paper applying the method to detect complex
        latent structures, such as replay in the striatum.

    \item\textbf{svGPFA extensions:} I am extending the core svGPFA framework
        to incorporate external latent drivers and real-time, online operation.

    \item\textbf{Bonsai.ML and Aeon data processing papers:} I am leading the
        writing of journal articles that introduce the \textit{Bonsai.ML}
        package for intelligent experimental control and, in collaboration with
        an exceptional undergraduate mentee, reporting novel statistical
        insights from the \textit{Aeon} project.

    \item\textbf{State-space models for behavioral inference:} To address the
        failure of deep pose-estimation networks to account for temporal
        continuity, we have developed
        \href{https://joacorapela.github.io/ssm/auto_examples/index.html#kinematics-inference}{state-space
        models to infer kinematics}.  I will describe these methods and
        demonstrate their advantages in capturing multi-animal behavioral
        continuity.

\end{itemize}

\paragraph{Evidence of potential to secure competitive research funding.}

During my postdoctoral position at UC San Diego I was awarded a seed grant from
\emph{Center for Brain Activity Mapping} and at the Gatsby Unit an £800.000
grant from BBSRC.

\paragraph{Evidence of effective supervision and mentoring of undergraduate,
postgraduate and/or doctoral research students including teaching in a
computational, data science, or quantitative context (e.g. programming,
machine learning, statistics, mathematical modelling) — ideally in neuroscience
or bioscience settings; Python experience strongly preferred.}

As part of the
\href{https://www.simonsfoundation.org/2021/09/09/inaugural-2021-scgb-undergraduate-research-fellowship-awardees-announced/}{Simons
Foundation's SURFiN program} I mentored an undergraduate student from QMUL and
another from UCL's NPP.

I have been teaching assistant, discussion leader, course coordinator and
instructor in courses in Python programming artificial intelligence, numerical
linear algebra, probability, statistical neuroscience, delivered to
undergraduate and graduate students. Please refer to my CV for details.

\paragraph{Evidence of productive cross-disciplinary collaborations,
speciﬁcally spanning computational or AI methods and neuroscience or biological
systems, for example joint A/I publications, grants, or collaborative projects
with researchers from complementary disciplines.}

I owe much of my research trajectory in computational neuroscience to close
partnerships with experimental collaborators:

\begin{itemize}

    \item During my PhD, I collaborated with Dr.~John Touryan and Dr.~Gidon
        Felsen (UC Berkeley, Yang Dan lab) to characterise responses to natural
        images in the cat primary visual cortex, using multivariate nonlinear
        regression models that I developed \citep{rapelaEtAl06,rapelaEtAl10}.

    \item During my first postdoc at UC San Diego, a collaboration with
        Prof.~Edward Chang (UCSF) helped me secure my first seed grant. His lab
        generously shared outstanding human ECoG recordings from epileptic
        subjects, which we used to fit and evaluate new ensemble dynamical
        models that we developed\citep{rapelaEtAlInPrepEDMs}.  These rich
        datasets ultimately enabled me to discover traveling waves in human
        language regions while subjects pronounced consonant-vowel syllables
        \citep{rapelaInPrepTWsInSpeech,rapelaInPrepSyncTWs,rapelaInPrepSyncTWsII}.

    \item Also, during my postdoc at UC San Diego, I collaborated with EEG
        experimental researchers, Prof.~Jeanne Townsend and Dr.~Marissa
        Westerfield on the application of investigation of the relation of
        single-trial phase coherence in the EEG and attentional shifts in
        humans~\citep{rapelaEtAl18-avshift}. The ML innovation of this research was the
        use of variational inference to estimate large-dimensional linear-regression
        models from limited amount of data.

    \item I was hired at the Gatsby Unit to collaborate with SWC experimental
        neuroscientists on the use of advanced statistical methods to analyse
        their recordings. One such successful collaboration with Ms.~ Mitra
        Javadsadeh No (Hofer lab) used linear dynamical system models, that I
        proposed and developed, to model long-range cortical communication. She
        presented orally this research at
        CoSyNe\footnote{\href{https://www.youtube.com/watch?v=lG1Rn4AwmY8}{Mitra
presented orally our results at CoSyNe 2022}. Presentation begins at 2:59:00.}

    \item At the SWC and Gatsby Unit I contributed my statistical expertise for
        the analysis of Aeon data~\citep{campagnerEtAl25b}. In this Aeon
        collaboration, I met Dr.~Goncalo Lopes, director of NeuroGEARS, with
        whom I applied, obtained and developed the BBSRC funded project
        \href{https://gtr.ukri.org/projects?ref=BB\%2FW019132\%2F1}{Machine
        Intelligence for Neuroscience Experimental Control}.

\end{itemize}

\paragraph{Experience of engagement beyond academia, including contributions to
policy, practice, public engagement, or knowledge exchange.}

My career has been deeply committed to knowledge exchange and transforming laboratory practice through the translation of complex machine learning into accessible, open-source scientific tools. Rather than keeping advanced mathematical frameworks confined within theoretical computing, I have focused on bridging the gap between machine learning and its practical, real-world application by experimental scientists.

\begin{itemize}
    \item \textbf{Knowledge Exchange and Community Building:} As a key co-developer of the \textit{Bonsai.ML} ecosystem and the statistical infrastructure for the institute-wide \textit{Aeon} project, I have directly facilitated knowledge exchange across distinct scientific domains. A prominent example of this engagement was my active involvement in the \textbf{2024 Bonsai Developers Conference}, where we brought together an international community of developers and experimentalists to share insights, standardize workflows, and lower the barrier to integrating machine learning into local laboratory environments. 
    \item \textbf{Transforming Practice via Open Software Distribution:} I have a proven, multi-year track record of developing and openly distributing robust statistical packages to the wider scientific community. By translating complex mathematical frameworks into production-ready software, I have directly enhanced computational workflows across laboratories. My publicly distributed contributions include:
    \begin{itemize}
        \item \href{https://github.com/joacorapela/svGPFA}{\textit{joacorapela/svGPFA}}: Python implementation for Sparse Variational Gaussian Process Factor Analysis.
        \item \href{https://github.com/joacorapela/ssm}{\textit{joacorapela/ssm}}: State-space models tailored for behavioral and kinematic inference.
        \item \href{https://github.com/joacorapela/hiddenMarkovModels}{\textit{joacorapela/hiddenMarkovModels}}: R implementations for HMMs with multivariate Gaussian, Poisson, Poisson GLM, and autoregressive observations.
        \item \href{https://github.com/joacorapela/bayesianLinearRegression}{\textit{joacorapela/bayesianLinearRegression}} \& \href{https://github.com/joacorapela/variationalLinearRegression}{\textit{joacorapela/variationalLinearRegression}}: Core computational frameworks for exact and variational Bayesian inference.
    \end{itemize}
    \item \textbf{Contributions to Educational Practice:} I have a proven track record of translating cutting-edge research into educational practice. I designed and taught the foundational \textit{Statistical Neuroscience} course for PhD students at the Sainsbury Wellcome Centre and mentored Simmons Foundation fellows. This work represents a sustained commitment to practice, ensuring that the next generation of biomedical and NeuroAI researchers are equipped with rigorous quantitative skills.
\end{itemize}
\subsection*{Skills and abilities}

\paragraph{Ability to develop and implement team-based solutions to complex challenges,
delivering positive outcomes.}

My entire career, from my PhD studies to my current role at UCL, has been built
upon the principles of team science. I firmly believe that the most complex
challenges in modern computational neuroscience cannot be solved in isolation;
they require the integration of diverse perspectives, technical skill sets, and
experimental methodologies. Rather than working as a solitary investigator, I
have consistently approached scientific obstacles by developing, leading, and
implementing team-based solutions alongside my collaborators.

\begin{itemize}

    \item \textbf{Bridging Theory and Experimentation:} Throughout my seven
        years at the Gatsby Computational Neuroscience Unit and the Sainsbury
        Wellcome Centre, I have served as the core collaborative bridge between
        advanced mathematical theory and complex biological experimentation.
        Working in continuous partnership with Unit Director Maneesh Sahani,
        experimental faculty, and engineering teams, I have co-developed
        institutional software infrastructure like \textit{Bonsai.ML} and the
        \textit{Aeon} project. These platforms succeeded precisely because they
        were designed as team-based solutions to the complex challenge of
        large-scale, automated behavioral and neural data acquisition.

    \item \textbf{Delivering Major Collaborative Outcomes:} This team-first
        approach has delivered verifiable, positive academic results. I was a
        primary driver in securing and executing an £800,000 BBSRC grant.
        Delivering on a grant of this scale required the continuous
        coordination of cross-disciplinary milestones, aligning
        data-engineering standards with the active scientific needs of multiple
        experimental laboratories.

    \item \textbf{A Distributed Mindset to Problem Solving:} From the beginning
        of my doctoral research to my recent open-source software
        distributions, every major technical challenge I have overcome has been
        solved through active, transparent engagement with my peers and
        mentors. I view software development, pedagogical design, and data
        analysis not as independent tasks, but as collaborative systems. 

\end{itemize}

As a faculty member, I intend to bring this established culture of team science
directly into the joint environment of NPP and CDB. I do not look to build an
isolated lab, but rather a collaborative node that actively partners with
experimentalists to solve complex neuroscientific challenges through shared,
team-based breakthroughs.

\paragraph{Advanced computational skills relevant to neuroscience research,
including proﬁciency in at least one scientiﬁc programming language (e.g.
Python, MATLAB, R, Julia) and familiarity with machine learning frameworks or
data analysis tools used in neuroscience or bioscience.}

I possess expert-level computational skills developed over more than a decade
of research and software engineering in computational neuroscience. I am highly
proficient in both Python and R for data analysis, statistical modeling, and
scientific visualization. Rather than merely being familiar with existing
machine learning frameworks, I actively build within them; I have engineered
and optimized the core architecture of the \textit{svGPFA} package utilizing
both PyTorch and JAX backbones to handle high-dimensional, real-world neural
population data.

\subsection*{UCL Ways of Working}

\paragraph{Ability to build collaborative research and industry partnerships.} 

The entirety of my research career in computational neuroscience provides
direct evidence of my ability to establish and sustain high-impact
collaborative research partnerships. My work naturally exists at the
intersection of diverse disciplines, requiring continuous synthesis between
theoretical modeling and experimental application. 

Beyond traditional academic networks, I have a proven track record of bridging
the gap between university research and commercial enterprise. A clear
testament to my ability to build industry partnerships is my successful
collaboration with \textbf{NeuroGEARS Ltd}. Together, we co-developed the
strategic vision, successfully bid for funding, and executed the development of
the \textit{Bonsai.ML} grant. This industry partnership has been vital in
translating advanced machine learning methods into robust, commercial-grade
software ecosystems that are optimized for active use in modern laboratories.
As a faculty member at UCL, I intend to leverage these established industry and
academic networks to drive innovative, well-funded cross-sector
collaborations.

\paragraph{Ability to develop and implement team-based solutions to complex challenges,
delivering positive outcomes.}

Answered above.

\paragraph{Active commitment to equality, diversity, and inclusion, for example
through inclusive teaching practices, mentoring underrepresented students,
participating in EDI initiatives, or fostering an inclusive lab/group
culture.}

Having been a member of an underrepresented minority in the US and the UK, I am
highly sensitive to EDI issues. This was one of my motivations for
participating in the SURFiN mentoring program of the Simons Foundation, where I
mentored a first student from Pakistan and a second one from China.

I am deeply dedicated to fostering an explicitly inclusive research and
classroom culture. In a faculty role at UCL, I will leverage my experiences to
implement equitable teaching practices in quantitative courses and actively
recruit, support, and advocate for diverse talent within the joint environment
of NPP and CDB.

\section*{Desirable Criteria}

\paragraph{Postgraduate Certiﬁcate in Teaching \& Learning in Higher Education,
or equivalent qualiﬁcation.}

While I do not currently hold a formal Postgraduate Certificate in Higher
Education, I am highly motivated to obtain this qualification and would
actively pursue it upon appointment. I have extensive practical experience in
educational design and delivery, having created and taught the foundational
\textit{Statistical Neuroscience} course for PhD students at the Sainsbury
Wellcome Centre and mentored international fellows. I view the opportunity to
complete a formal teaching certificate at UCL as a valuable mechanism to ground
my existing pedagogical experience in structured higher-education theory and to
ensure my classroom methods remain inclusive and highly effective.

\paragraph{Experience of contributions to peer review bodies/committees, professional
organisations, learned societies, government committees or Research Councils etc.}

I am an ad-hoc reviewer for Vision Research, Frontiers in Neuroscience, and Journal
of Perceptual Imaging. In addition, I have reviewed grant proposal for the
Software Sustainability Institute.

\paragraph{Experience of an active role in widening participation initiatives
that span research communities, e.g. sharing ideas and experience through
organising seminars and workshops.}

I believe that advancing scientific discovery requires building open,
transparent, and equitable research environments. To actively contribute to
this mission, I served as a member of the Sainsbury Wellcome Centre
\href{https://sainsburywellcomecentre.github.io/RCWG/}{Research Culture Working
Group (RCWG)}.

In this role, I collaborated with a cross-disciplinary team of researchers and
staff to evaluate and systematically improve the research culture of our
institute. My contributions focused on fostering a highly collaborative
environment that explicitly bridges the gap between theoretical modeling and
experimental laboratories. By designing initiatives that encourage the open
sharing of data, software workflows, and technical expertise across distinct
scientific communities, this work directly advanced widening participation,
knowledge equity, and intellectual accessibility within the broader SWC and
Gatsby Unit ecosystem.

\pagebreak
\bibliographystyle{apalike}
\bibliography{latentVariableModels,replay,longDurationExperimentation,receptiveFields,eeg,mesoscaleModels,travelingWaves,other}

\end{document}
