\documentclass{article}

\usepackage[margin=1.5cm]{geometry}
\usepackage[colorlinks]{hyperref}
\usepackage{natbib}

\title{Vision Statement}
\author{Joaquín Rapela}

\begin{document}

% \maketitle

\section*{Research Vision}

\textit{Mathematics without biology or biology without mathematics are not as
wonderful}.

% Since pursuing my PhD, my research career has been deeply interdisciplinary
% spanning analytical (e.g., mathematics, signal processing, statistics, machine
% learning, computer science) and experimental neuroscience subjects (e.g.,
% statistical analysis of responses of single cells, electrocorticography, Utah
% arrays, Neuropixels recordings, close-loop experimental control). I delight
% when I create a new method to process biological data, but I delight further
% when I can apply it to discover new aspects of brain function.

My \emph{NeuroAI Research Group} will initially focus on
creating advanced statistical methods to characterise neural and behavioral
data, offline and in real time, for short- and long-duration experiments. In
addition, in collaboration with experimental groups, we will apply these
methods, or those created by others, to investigate behavioural and neural
questions.
%
% The research line that my group will follow best aligns with those of
% \href{https://sites.stat.columbia.edu/liam/}{Prof.~Liam Paninski} and
% \href{https://pillowlab.princeton.edu/}{Prof.~Jonathan Pillow}.

This is an exciting time to work on NeuroAI. As the variety of
neural datasets being recorded (e.g., high-channel-count electrophysiology, one- and
two-photon calcium imaging, voltage imaging, EM, transcriptonics, connectomics)
keeps increasing, and their size and complexity become larger and larger, due to
more naturalistic and longer experimentation, there is a critical need for
statistical methods that can draw efficient inferences and extract meaning from
these datasets.
%
In addition, as new optogenetic methods are developed to spatio-temporally
manipulate neural activity, we need these inferences in real time, combined
with methods from optimal control, to maximise the efficiency of these
manipulations.
%
Yet, we are currently not seeing an increase in the number of NeuroAI
scientists,
trained in both statistical machine learning and neurosciences, required to
optimally process these datasets with advanced statistical ML methods.
%
My research group will attempt to reverse this trend.

\paragraph{Latent variable models}

It is challenging to extract meaning from the large number of spikes generated
by high-channel-count recordings probes, like Neuropixels. To address this
problem, my supervisor at the Gatsby Unit, Prof.~Maneesh Sahani, and
collaborators created
Gaussian Process Factor Analysis~\citep[GPFA;][]{yuEtAl09}, a probabilistic
method to extract low-dimensional representations (i.e., latent variables) from
binned spike-counts in high-dimensional electrophysiological recordings. He recently extended it in
Sparse Variational Gaussian Process Factor Analysis
\citep[svGPFA;][]{dunckerAndSahani18} with sparse-variational methods to
substantially reduce
the computational and memory requirement of GPFA, allowing svGPFA to
characterise longer trials than GPFA, and allowing point-process observations,
eliminating the need to bin spikes times in GPFA.
%
At the Unit I have distributed a highly optimized Python implementation of
svGPFA and explored its capabilities to
\href{https://joacorapela.github.io/svGPFA/auto_examples/index.html}{characterize
several high-channel count neural datasets}.

In collaboration with the laboratory of Marcus Stephenson-Jones at the SWC, I
have extracted svGPFA neural latents from striatal Neuropixels recordings in
mice performing a stereotyped poking sequence task~\citep{thompsonEtAl24}. I
then used these neural latents as inputs to a Hidden Markov Model to decode
mice behavioral states with remarkably high accuracy. We are now observing that
patterns of neural latents detected during behavior replay during sleep.
Similar decoding and replay are obtained using latent variables estimated using
linear dynamical systems, instead of svGPFA.
%
If these observations are confirmed, it will be the first time that accurate
decoding is achieved and replay is detected in continuous latent variables of
large neural populations, bypassing the estimation of place fields.

My group will continue developing latent variable models for neural populations
and will collaborate with experimental groups at NPP, CDB and elsewhere, and
use them to characterize neural datasets and understand brain function. I am
specifically interested in using these models for position decoding and replay
detection in the hippocampus.

\paragraph{Latent variable models} 

Extracting meaningful population dynamics from high-channel-count recording
probes, such as Neuropixels, remains a major computational challenge. To
address this, Gaussian Process Factor Analysis~\citep[GPFA;][]{yuEtAl09} was
introduced to extract low-dimensional latent variables from population
activity. This framework was subsequently extended into Sparse Variational
GPFA~\citep[svGPFA;][]{dunckerAndSahani18} using sparse-variational methods to
drastically reduce computational and memory overhead. Crucially, svGPFA
accommodates direct point-process observations, eliminating the need to bin
spike times and enabling the characterisation of significantly longer
experimental trials.

As the lead developer of this ecosystem, I built and distributed a highly
optimised Python implementation of svGPFA, leveraging it to characterise
several large-scale neural datasets. In collaboration with the Stephenson-Jones
laboratory at the SWC, I extracted svGPFA neural latents from striatal
Neuropixels recordings in mice performing a stereotyped poking sequence
task~\citep{thompsonEtAl24}. By utilizing these latents as inputs to a Hidden
Markov Model (HMM), I decoded fine-grained behavioural states with exceptional
accuracy. Strikingly, we are now discovering that these low-dimensional neural
trajectories replay during sleep. While similar structural dynamics can be
unmasked using standard linear dynamical systems (LDS), svGPFA's unbinned
point-process formulation uniquely preserves fine-grained temporal features.
These results mark a major milestone: the first demonstration of accurate state
decoding and replay tracking within the continuous latent variables of large
neural populations, entirely bypassing the traditional requirement of
estimating place fields.

Moving forward, my group will pioneer next-generation latent variable models
tailored for massive neural populations. Partnering with experimental groups
across NPP, CDB, and the wider UCL community, we will deploy these tools to map
hidden brain dynamics and understand systemic circuit function. A core
objective of my independent research program will be applying these
probabilistic frameworks to resolve open questions surrounding non-spatial
position decoding and replay detection in the hippocampus.

\paragraph{Latent variable models}

Extracting meaningful structure from the massive spike trains generated by
high-channel-count probes like Neuropixels is notoriously challenging. To
address this, Gaussian Process Factor Analysis~\citep[GPFA;][]{yuEtAl09} was
developed to extract low-dimensional latent variables from binned spike counts.
This approach was recently extended into Sparse Variational
GPFA~\citep[svGPFA;][]{dunckerAndSahani18} to substantially reduce
computational and memory bottlenecks. By incorporating point-process
observations, svGPFA eliminates heuristic spike binning entirely and
successfully characterises much longer experimental trials than standard GPFA.

At the Gatsby Unit, I developed and distributed a highly optimised Python
implementation of svGPFA to explore its capabilities across several
high-channel neural datasets. In collaboration with the Stephenson-Jones lab at
the SWC, I applied this framework to striatal Neuropixels recordings from mice
executing a stereotyped poking sequence task~\citep{thompsonEtAl24}. By feeding
these neural latents into a Hidden Markov Model, I decoded distinct behavioural
states with high fidelity. We have now observed that these latent neural
patterns actively replay during sleep. Notably, while comparable decoding and
replay profiles are obtainable via linear dynamical systems, svGPFA offers
unique advantages due to its continuous-time point-process foundation. Once
confirmed, these findings will represent the first time that accurate decoding
and replay tracking have been achieved directly within the continuous latent
space of large populations, bypassing the need to estimate explicit place
fields.

My independent research group will continue to develop advanced latent variable
models for population recordings. We will collaborate closely with experimental
labs at NPP, CDB, and internationally to characterise novel datasets and map
fundamental brain functions. Specifically, my lab will focus on adapting these
probabilistic models to investigate unbinned position decoding and replay
dynamics within hippocampal circuits.

\paragraph{NeuroAI methods for naturalistic, long-duration and continual
experimentation}

At the SWC and Gatsby Unit, we are pioneering a new type of naturalistic,
long-duration and continual experimentation, implemented in the
\href{https://aeon.swc.ucl.ac.uk/}{Aeon platform}, that promises to
revolutionise neuroscience experimentation~\citep{campagnerEtAl25b}.

In 2023-2024 I mentored a very talented undergraduate student from NPP,
\href{https://zimoli02.github.io/}{Ms.~Zimo Li}, in a
\href{https://www.simonsfoundation.org/2023/09/26/simons-foundation-announces-next-class-of-neuroscience-undergraduate-research-fellows/}{Simons
Foundation SURFiN
fellowship}, where she performed the first comprehensive statistical analysis
of Aeon data.
%
The achievements of Zimo were outstanding. She processed very complex data,
that at that stage were poorly organised, and skillfully applied sophisticated
statistical models (e.g., linear dynamical systems, hidden Markov models, point
process regression) and produced the first statistical Aeon discoveries, which
were included in the
\href{https://www.biorxiv.org/content/10.1101/2025.07.31.664513v1}{Aeon
platform paper}, with Zimo as coauthor. The collaboration with Zimo taught me
that I can do excellent research with students that do not have a solid
mathematical background.

Due to the large size of Aeon recordings, conventional batch machine learning
(ML) methods cannot be used to process Aeon data. Aeon requires online methods
that are able to process non-stationary data and are computationally very
efficient to handle very large datasets. A detailed description of my ML data
processing plans for Aeon data appears in a
\href{https://www.gatsby.ucl.ac.uk/~rapela/npp/submittedProposal.pdf}{collaborative
BBSRC/NSF grant we submitted jointly with the Allen Institute for Neural
Dynamics in 2024}.  My group will continue contributing advanced statistical
methods for the characterisation of Aeon data offline and in real time..

\paragraph{NeuroAI methods for intelligent experimental control}

Current neuroscience experimentation is mostly performed in open loop with
predefined sets of rules designed by experimenters. More informative
experiments can be designed if they could be controlled in close loop by
intelligent inferences performed on behavioural and neural data streams.

\href{https://bonsai-rx.org/}{Bonsai} is today the most widely used software
for close loop experimental control in neuroscience. In 2022 I realised that
adding ML functionality to Bonsai could make possible a radically new type of
close loop intelligent experimentation, and led the successful application of
a BBSRC proposal
(\href{https://gtr.ukri.org/projects?ref=BB\%2FW019132\%2F1}{BB/W019132/1}) to
build the \href{https://bonsai-rx.org/machinelearning/}{Bonsai.ML package}.

Most current ML methods operate offline, trained and performing inference on
datasets previously recorded. However, Bonsai.ML requires a different type of ML
methods, which are trained and perform inference on live streams, continuously
delivering data in real time. We built this type of ML into Bonsai.ML. It now
contains probabilistic ML methods for real time processing of behavioral (e.g.,
kinematics inference with linear dynamical systems, behavioral state estimation
with hidden Markov models) and neural data (neural latent variables inference
with linear dynamical systems, hippocampal position decoding and replay detection with a
point process state-space model). Please refer to the
\href{https://bonsai-rx.org/machinelearning/}{documentation of
Bonsai.ML}.

We are now integrating large language models (LLMs) into Bonsai. We are using
them to help Bonsai users build correct and optimised workflows using natural
language. Additionally, we are building a Bonsai LLM node to bring the large
semantic knowledge of LLMs into the experimental controls loop.  Please refer
to the \href{https://github.com/joacorapela/os4ls}{Open Source for the Life
Science proposal} we are currently preparing to support these developments.

My group will continue developing real-time probabilistic ML models to enable
intelligent experimental control in Bonsai.

\paragraph{Collaborations}

I owe much of my research career in computational neuroscience to my
experimental collaborators.
%
While doing my PhD, I worked with John Touryan and Gidon Felsen, both at UC
Berkeley,  on the analysis of their recordings from visual cells in the cat
primary visual cortex from their responses to natural images with a method that
I developed to estimate multivariate nonlinear regression
models~\citep{rapelaEtAl06,rapelaEtAl10}.
%
During my first postdoc at UC San Diego, I obtained my first seed grant thanks
to the generosity of my collaborator Edward Chang, UC San Francisco, that
shared with us outstanding ECoG recordings from epileptic subjects to model
with ensembles dynamical models~\citep{rapelaEtAlInPrepEDMs}. Later these
recordings allowed me to find traveling waves in speech brain regions while
humans pronounced consonant-vowel
syllables~\citep{rapelaInPrepTWsInSpeech,rapelaInPrepSyncTWs,rapelaInPrepSyncTWsII}.
%
And, although I don't have space to mention them, all collaborators have played
a critical role in my career.

In 2019 the Gatsby Unit hired me to facilitate collaborations between the Unit
and the SWC.
%
At the SWC experimental scientists collect rich and complex neural and
behavioral datasets. Because often they don't have the analytical skills to best
characterise these recordings, they end up using simple and sub optimal
statistical techniques.
%
I worked with them to find good statistical methods to model their data.
Below I highlight one of these collaborations.

To explore collaborative opportunities I attended the SWC Data Clubs, where SWC
scientists presented their latest findings. In one data club Ms.~Mitra
Javadzadeh, Ph.D.\ student in Prof.~Hofer's lab, presented her long-range
cortical communications research, where she
was using Linear Discriminant Analysis to model this communication. I proposed
to use linear dynamical systems instead, as the dynamic characteristic of
this model better suited her experimental setup. Mitra shared with me a sample
of her data and asked me to try my suggestion on it. COVID time was ideal for
this analysis, which produced very interesting results. I shared these initial
results with her, her advisor and my supervisor joined the investigation, and
\href{https://proceedings.neurips.cc/paper_files/paper/2021/file/859b755563f548d008f936906a959c8f-Paper.pdf}{Mitra
presented orally our results at CoSyNe 2022}\footnote{Mitra's presentation starts at
2:59:00}. Now, as faculty member at CSHL, Mitra continues this line of
research~\citep{javadzadehEtAl24}.
%
My collaboration my Mitra, as well as those with other SWC scientists, taught
me that methodological/experimental collaborations have long-term effect on the
research and the carrers of my collaborators.

I am currently writing an article for
\href{https://www.thetransmitter.org/}{The Trasmitter} discussing the need for
more NeuroAI scientists in Neuroscience. As I outlined at the beginning, we
need many more NeuroAI scientists, trained in both statistical machine learning
and neuroscience, to help experimental neuroscientists extract more meaning
from their rich recordings.
%
I want to create a small team of NeuroAI scientists that will collaborate with
experimental neuroscientists to optimally process their datasets.

\section*{Teaching Vision}

I am experienced teaching quantitative subjects to life sciences students.
Please refer to the Teaching section of my CV.

My teaching philosophy has been heavily influenced by my postdoctoral mentor,
Prof.~Wilson Truccolo, Brown University, and aligns closely with the style of
\href{https://mark-kramer.github.io/Case-Studies-Python/intro.html}{Case
Studies in Neural Data Analysis}, by Prof.~Mark Kramer, Boston University.
%
I deliver lectures that are mathematically rigorous, but skip proofs from the
main thread of content. These proofs are provided in appendices.
%
I complement my lectures with many examples, most often in the form of computer
code. These examples help students grasp the utility of the concepts I teach,
even if they don't fully understand their mathematical underpinning.
%
I design my lectures to be multidimensional, so that both naive and expert
students can benefit from them.
%
Please refer to the
\href{https://github.com/joacorapela/gcnuBridging2023/blob/master/probability/lectures/2_foundations/foundations.pdf}{Foundations
of probability  theory} lecture, delivered in the first edition of Gatsby
Unit's Bridging Program in 2023, as a demonstration of my teaching style.

I am very enthusiastic about learning, and from students' feedback I learned
that some of this enthusiasm transpires in my teaching.

\pagebreak
\bibliographystyle{apalike}
\bibliography{latentVariableModels,replay,longDurationExperimentation,receptiveFields,mesoscaleModels,travelingWaves,other}

\end{document}
